---
title: "Project_1"
author: "Muhammad Enrizky Brillian_1009713712"
date: '2023-04-04'
output: html_document
---

# Merging Datasets from multiple CSVs
```{r, eval=F}
# loading libraries
library(tidyverse)
library(dplyr)
#Setting the working Directory
setwd("~/STAA57_Project")
# Create empty data frame
bike_data <- data.frame()

# Loop through all CSV files in folder
for (file in list.files(pattern=".csv")) {
  temp_data <- read.csv(file, header=TRUE)  # Read in CSV file
  bike_data <- rbind(bike_data, temp_data)  # Bind data to existing data frame
  rm(temp_data)  # Remove original data frame to free up memory
}

```

# Description of the Dataset
```{r}
names(bike_data)
```

1. Trip ID: A unique identifier for each trip
2. Trip Duration (Seconds): The length of the trip in seconds
3. From Station ID: The unique identifier for the starting station of the trip
4. Trip Start Time: The date and time the trip began
5. From Station Name: The name of the starting station for the trip
6. Trip Stop Time: The date and time the trip ended
7. To Station ID: The unique identifier for the ending station of the trip
8. Bike ID: The name of the ending station for the trip
9. User Type: A categorical variable indicating whether the user is a "Annual Member" (annual pass holder of the bikeshare program) or a "Casual Member"(24 or 72 hour pass holders of the bikeshare program)



